---
title: "Sensivity Analysis - Interest Rate vs EPV"
author: "Sivaram Ainkaran"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tinytex)
library(tidyverse)


```

```{r, echo=FALSE}
Data <- read.table('mortality_table.txt', dec='.',sep='\t', header=TRUE)
#read the l_x table into a usable table in R
Data
```

```{r, echo=FALSE}
interest <- seq(0,0.5,0.01) #Vector of length 50 for interest rate
a_0 <- c()  #Empty vector a_0 values are added to for each interest rate
a_30 <- c() #Empty vector a_30 values are added to for each interest rate
a_60 <- c() #Empty vector a_60 values are added to for each interest rate
A_0 <- c()  #Empty vector A_0 values are added to for each interest rate
A_30 <- c() #Empty vector A_30 values are added to for each interest rate
A_60 <- c() #Empty vector A_60 values are added to for each interest rate
```


```{r, echo=FALSE}
#'a_0'
g=1                     #age of participant + 1
for (i in 0:50){        #for loop to generate a_0 for each interest rate
  v <- 1/(1+(i/100))    #generate v to input to each a_0
  ax = c()              #create empty vector used to calculate a_0
  for (k in 1:104){     #for loop to generate a_0 for given interest rate
    lx <- Data$lx[g]    #value of l_x at age x ie. g-1 from given Data
    lx1 <- Data$lx[k+g] #value of l_(x+k) at age x ie. g-1 from given Data
    if (k+g<=104){      #if statement to ensure no errors when k+g>104 when calculating kpx
      kpx <- lx1/lx     #value of kpx is l_(x+k)/l_x
    }else{
      kpx <- 0          #if k+g>104, kpx=0, to prevent any errors
    }
  a_x <- (v^k)*kpx      
  ax[k] <- a_x          #store values in ax vector
  }
  a0 <- sum(ax)         #sum values in ax vector to find a_0 for given i
a_0[i+1] <- a0          #store each of these into the a_0 vector 
}
a_0 <- a_0/a_0[1]       #standardise each a0 value in the a_0 vector




#'A_0'
for (i in 0:50){         #for loop to generate A_0 for each interest rate
  v <- 1/(1+(i/100))     #generate v to input to each A_0
  ax = c()               #create empty vector used to calculate A_0
  for (k in 0:104){      #for loop to generate A_0 for given interest rate
    lx <- Data$lx[g]     #value of l_x at age x ie. g-1 from given Data
    lx1 <- Data$lx[k+g]  #value of l_(x+k) at age x ie. g-1 from given Data
    lqx <- Data$lx[k+g+1]#value of l_(x+k+1) at age x ie. g-1 from given Data
    if (k+g<=104){       #if statement to ensure no errors when k+g>104 when calculating kpx
      kpx <- lx1/lx      #value of kpx is l_(x+k)/l_x
    }else{
      kpx <- 0           #if k+g>104, kpx=0, to prevent any errors
    }
    if (k+g+1<=104){     #if statement to ensure no errors when k+g+1>104 when calculating (k+1)px
      kqx <- lqx/lx      #value of (k+1)px is l_(x+k+1)/l_x
    }else{
      kqx <- 0           #if k+g+1>104, (k+1)px=0, to prevent any errors
    }
  k_qx <- kpx-kqx        #kqx is kpx-(k+1)px
  a_x <- (v^(k+1))*k_qx  
  ax[k+1] <- a_x         #store values in ax vector
  }
  A0 <- sum(ax)          #sum values in ax vector to find A_0 for given i
A_0[i+1] <- A0           #store each of these into the A_0 vector 
}
A_0 <- A_0/A_0[1]        #standardise each a0 value in the A_0 vector
```

```{r, echo=FALSE}
#comments are similar for all ages since procedures are the same
#'a_30'
g=31
for (i in 0:50){
  v <- 1/(1+(i/100))
  ax = c()
  for (k in 1:104){
    lx <- Data$lx[g]
    lx1 <- Data$lx[k+g]
    if (k+g<=104){
      kpx <- lx1/lx
    }else{
      kpx <- 0
    }
  a_x <- (v^k)*kpx 
  ax[k] <- a_x
  }
  a30 <- sum(ax)
a_30[i+1] <- a30
}
a_30 <- a_30/a_30[1]




#'A_30'
for (i in 0:50){
  v <- 1/(1+(i/100))
  ax = c()
  for (k in 0:104){
    lx <- Data$lx[g]
    lx1 <- Data$lx[k+g]
    lqx <- Data$lx[k+g+1]
    if (k+g<=104){
      kpx <- lx1/lx
    }else{
      kpx <- 0
    }
    if (k+g+1<=104){
      kqx <- lqx/lx
    }else{
      kqx <- 0
    }
  k_qx <- kpx-kqx
  a_x <- (v^(k+1))*k_qx 
  ax[k+1] <- a_x
  }
  A30 <- sum(ax)
A_30[i+1] <- A30
}
A_30 <- A_30/A_30[1]

```



```{r, echo=FALSE}
#'a_60'
g=61
for (i in 0:50){
  v <- 1/(1+(i/100))
  ax = c()
  for (k in 1:104){
    lx <- Data$lx[g]
    lx1 <- Data$lx[k+g]
    if (k+g<=104){
      kpx <- lx1/lx
    }else{
      kpx <- 0
    }
  a_x <- (v^k)*kpx 
  ax[k] <- a_x
  }
  a60 <- sum(ax)
a_60[i+1] <- a60
}
a_60 <- a_60/a_60[1]




#'A_60'
for (i in 0:50){
  v <- 1/(1+(i/100))
  ax = c()
  for (k in 0:104){
    lx <- Data$lx[g]
    lx1 <- Data$lx[k+g]
    lqx <- Data$lx[k+g+1]
    if (k+g<=104){
      kpx <- lx1/lx
    }else{
      kpx <- 0
    }
    if (k+g+1<=104){
      kqx <- lqx/lx
    }else{
      kqx <- 0
    }
  k_qx <- kpx-kqx
  a_x <- (v^(k+1))*k_qx 
  ax[k+1] <- a_x
  }
  A60 <- sum(ax)
A_60[i+1] <- A60
}
A_60 <- A_60/A_60[1]

```


```{r, echo=FALSE}

par(bg='white',fg='black', font.axis=2)
df <- data.frame(interest,a_0,a_30,a_60,A_0,A_30,A_60)
write.table(df,'29741823_analysis_output.txt',sep='\t',dec='.',row.names=FALSE)
plot(a_0,type="l", xlab="Interest rate (%)",col.axis='black',col.lab='black',ylab="Standardised EPV",main="Interest rate's effect on EPV")
lines(a_30,type="l",col="blue")
lines(a_60,type="l",col="green")
lines(A_0,type="l",col="orange")
lines(A_30,type="l",col="purple")
lines(A_60,type="l",col="black")
legend("topright",legend=c("Annuity_0","Annuity_30","Annuity_60","Assurance_0","Assurance_30","Assurance_60"), col=c("red","blue","green","orange","purple","black"),lty=1,cex=0.8,text.font=4,text.col="black",bg="aliceblue")
```


The whole life annuity product, when compared with the whole life assurance product tends to have a higher standardised EPV (Expected Present Value) over all interest rates at each respective age ie. at x = 0, 30, 60.
This is due to $_{k}p_{x}$ having a higher value than $_{k|}q_{x}$ at all interest rates and ages since $_{k|}q_{x} =\ _{k}p_{x} -\ _{k+1}p_{x}$.


We can see that as age(x) decreases, both the $a_{x}$ and $A_{x}$ curves tend to drop more quickly and seem to plateau. This is due to the fact that at lower ages there are much more values used to calculate $a_{x}$ and $A_{x}$ since there are more ages between them and the maximum age. This means for the low interest rates all these values will contribute towards computing $a_{x}$ and $A_{x}$ but as $i$ increases and $k$ increases the discount rate will tend towards 0 meaning values as k increases make less and less of a contribution in calculating $a_{x}$ and $A_{x}$ and causes a rapid dip in these values as age increases.

The product standardised expected present values, $A_{0}, A_{30}\ and \ a_{0}$ all are very sensitive to interest rate changes and plateau very quickly, dropping below a standardised EPV of 0.2 by interest rates of 4%, 7% and 9% respectively.The standardised expected present values $a_{30}, A_{60}\ and \ a_{60}$ on the other hand are not as sensitive and do not plateau as fast, dropping below a standardised EPV of 0.2 by interest rates of 16%, 25% and 43% respectively. 



